### 2016-05-01 Paper_v5.tex
Relase candidate

Meta-heuristics can be effective techniques to estimate optimal or nearly optimal solutions for complex, multi-modal functions.
However, meta-heuristics typically require a large number of function evaluations to converge to a solution.
DOPS is a combination of particle swarm optimization, which is a global search method, and dynamically dimensioned search, which is a greedy evolutionary technique.
Particle swarm optimization uses collective information shared amongst swarms of computational particles to
search for global extrema. Several particle swarm variants have been proposed to improve the search ability and rate of convergence.
These variations involve different neighborhood structures, multi-swarms or adaptive parameters.
Multi-swarm particle swarm optimization with small particle neighborhoods has been shown to be better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
Multi swarm methods generate diverse solutions, and avoid rapid convergence to local optima.
However, at least for the coagulation problem used in this study, multi-swarm methods stagnated after approximately 25\% of the available function evaluations; only the introduction of dynamically dimensioned search improved the rate of error convergence.
Dynamically dimensioned search, which greedily perturbs only a subset of parameter dimensions in high dimensional parameter spaces,
refined the globally best particle and produced significantly lower error on average when compared to multi-swarm particle swarm optimization alone.
However, dynamically dimensioned search, starting from a initial random parameter guess, was not as effective on average as DOPS.
The initial solutions generated by the multi swarm search had a higher propensity to produce good parameter estimates when refined by dynamically dimensioned search.
Thus, our hybrid combination of two meta-heuristics produced better results than either constituent approach, and better results than other meta-heuristic approaches on average.
This was true of not only the convergence rate on the coagulation problem, but also the biochemical benchmark problems;
DOPS required two-orders of magnitude fewer function evaluations compared with enhanced Scatter Search (eSS) to estimate the biochemical benchmark model parameters.
Taken together, the combination of particle swarm optimization and dynamically dimensioned search produced better results than either of these constituent approaches alone,
and required fewer function evaluations compared with other common meta-heuristics.

%to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
%In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
%Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions  is an effective way to obtain near optimal solutions with few function evaluations.
%We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search. Thus  Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone. The computational cost was prohibitively high for parameter estimation on this problem using a standard PSO. Thus we used SA, DE and DDS for the purpose of comparison. Choosing the number of function evaluations is largely a function of cost and complexity of the objective function. Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold. However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion. In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case. We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}. In this work by Villaverde et al. they used a population based search enhanced Scatter Search (eSS) to estimate the biochemical model parameters. Quite surprisingly we took a couple of orders lesser number of function evaluations 10\textsuperscript{3} to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 10\textsuperscript{5} number of evaluations.  The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
%A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem. In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase. We used the same parameters for all the problems. The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing. For SA, we optimized the cooling schedule for the coagulation model. Thus, in this approach any overhead that usually comes with additional function evaluations to pre optimize parameters was avoided.

The performance of DOPS was impressive given that it performed well on different complex systems with no pre optimization of algorithm parameters being required. We comfortably outperformed existing, widely used meta-heuristics and were also able to find minima of high dimensional global optimization test functions. Thus this approach may be well suited to large scale global optimization. In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared to enhanced Scatter Search (eSS). However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems. This aspect is currently beyond the scope of this study. Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions. In addition, the current implementation of the algorithm is designed to switch only once from the swarm phase to the DDS. In the DDS phase, the search uses only one candidate vector although there is a provision to start with a population of candidate vectors. Incorporating a more intelligent switching strategy that can do switch from swarm phase to DDS phase multiple times and having a population of candidate vectors are some aspects of the algorithm that can be studied further.

>>>>

Meta-heuristics can be effective techniques to estimate optimal or nearly optimal solutions for complex, multi-modal functions.
However, meta-heuristics typically require a large number of function evaluations to converge to a solution.
DOPS is a combination of particle swarm optimization, which is a global search method, and dynamically dimensioned search, which is a greedy evolutionary technique.
Particle swarm optimization uses collective information shared amongst swarms of computational particles to
search for global extrema. Several particle swarm variants have been proposed to improve the search ability and rate of convergence.
These variations involve different neighborhood structures, multi-swarms or adaptive parameters.
Multi-swarm particle swarm optimization with small particle neighborhoods has been shown to be better in searching on complex multi-modal solutions \cite{zhao2008dynamic}.
Multi-swarm methods generate diverse solutions, and avoid rapid convergence to local optima.
However, at least for the coagulation problem used in this study, multi-swarm methods stagnated after approximately 25\% of the available function evaluations; only the introduction of dynamically dimensioned search improved the rate of error convergence.
Dynamically dimensioned search, which greedily perturbs only a subset of parameter dimensions in high dimensional parameter spaces,
refined the globally best particle and produced significantly lower error on average when compared to multi-swarm particle swarm optimization alone.
However, dynamically dimensioned search, starting from a initial random parameter guess, was not as effective on average as DOPS.
The initial solutions generated by the multi swarm search had a higher propensity to produce good parameter estimates when refined by dynamically dimensioned search.
Thus, our hybrid combination of two meta-heuristics produced better results than either constituent approach, and better results than other meta-heuristic approaches on average.
This was true of not only the convergence rate on the coagulation problem, but also the biochemical benchmark problems;
DOPS required two-orders of magnitude fewer function evaluations compared with enhanced Scatter Search (eSS) to estimate the biochemical benchmark model parameters.
Taken together, the combination of particle swarm optimization and dynamically dimensioned search produced better results than either of these constituent approaches alone,
and required fewer function evaluations compared with other common meta-heuristics.

%to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution.
%In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization.
%Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions  is an effective way to obtain near optimal solutions with few function evaluations.
%We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search. Thus  Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone. The computational cost was prohibitively high for parameter estimation on this problem using a standard PSO. Thus we used SA, DE and DDS for the purpose of comparison. Choosing the number of function evaluations is largely a function of cost and complexity of the objective function. Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold. However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion. In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case. We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}. In this work by Villaverde et al. they used a population based search enhanced Scatter Search (eSS) to estimate the biochemical model parameters. Quite surprisingly we took a couple of orders lesser number of function evaluations 10\textsuperscript{3} to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 10\textsuperscript{5} number of evaluations.  The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
%A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem. In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase. We used the same parameters for all the problems. The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing. For SA, we optimized the cooling schedule for the coagulation model. Thus, in this approach any overhead that usually comes with additional function evaluations to pre optimize parameters was avoided.
%In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared to enhanced Scatter Search (eSS).
%In addition, In the DDS phase, the search uses only one candidate vector although there is a provision to start with a population of candidate vectors. Incorporating a more intelligent switching strategy that can do switch from swarm phase to DDS phase multiple times and having a population of candidate vectors are some aspects of the algorithm that can be studied further.

DOPS performed well on many different systems with no pre-optimization of algorithm parameters, however there are many research questions that should be pursued further.
DOPS comfortably outperformed existing, widely used meta-heuristics for high dimensional global optimization functions, biochemical benchmark models and a model of the human coagulation system.
However, it is possible that highly optimized versions of common meta-heuristics could surpass DOPS;
we should compare the performance of DOPS with optimized versions of the common
meta-heuristics on both test and real-world problems to determine if a performance advantage exists in practice.
Next, DOPS has a hybrid architecture, thus the particle swarm phase could be combined with other search strategies such as local derivative based approaches to improve convergence rates.
We could also consider multiple phases beyond particle swarm and dynamically dimensioned search, for example switching to a gradient based search following the dynamically
dimensioned search phase.
Next, the current implementation of the algorithm switches only once from the swarm phase to the dynamically dimensioned phase.
An obvious modification would be to allow multiple transitions between the swarm and dynamically dimensioned phases.
Lastly, we should update DOPS to treat multi-objective problems.
The identification of large biochemical models sometimes requires training using qualitative, conflicting or even contradictory data sets.
One strategy to address this challenge is to estimate experimentally constrained model ensembles using multiobjective optimization.
Previously, we developed Pareto Optimal Ensemble Techniques (POETs) which integrates simulated annealing with Pareto optimality
to identify models near the optimal tradeoff surface between competing training objectives \cite{song2010ensembles}.
Since DOPS consistently outperformed simulated annealing on both test and real-world problems, we expect a multi-objective form of DOPS would more quickly estimate solutions
which lie along high dimensional trade-off surfaces.
